{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming\\[0.25em]

### 2025-07-13 paper.tex
Camera ready

\newcommand{\diff}[1]{{\textcolor{magenta}{#1}}}
\newcommand{\note}[1]{\marginpar{\raggedright\scriptsize\textcolor{magenta}{#1}}}

%\newcommand{\diff}[1]{{#1}}
%\newcommand{\note}[1]{}

@@

\begin{teaserfigure}
  \vspace{-0.2em}
  \includegraphics[width=1\textwidth,clip,trim=0cm 8.5cm 7.7cm 0cm]{fig/teaser.pdf}
  \vspace{-1.5em}
  \caption{\diff{Denicek is a computational substrate for document-oriented programming
    based on document edit histories~(A). We co-design Denicek with a web-based end-user
    programming environment Denicek (B). Here, Denicek is used to build a Todo list app via
    programming-by-demonstration, by copying a value from input to a new list item.
    We then evaluate the generality of Denicek by using it to build a data science notebook
    system Datnicek (C). Here, Datnicek is used to interactively clean the data table, by
    removing the `` p'' (provisional) marker from numerical columns.
  }}
  \Description{The Denicek diagram in the middle illustrates a document representation as an
   edit history. Two arrows point from the diagram to two screenshots. Webnicek on the left
   contains a Todo list app with a command palette showing add item operation. Datnicek on the
   right contains two cells, one with code to load data and another with an interactive editor
   for data cleaning.}
  \label{fig:teaser}
  \vspace{0.75em}
\end{teaserfigure}
\maketitle

@@

\begin{itemize}
\item \emph{Collaborative editing.} Users should be able to locally modify a shared document
  and merge their changes, preferably in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  enacting the steps of the desired behavior using concrete examples and generalizing from the example~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on the part are invalidated and, possibly automatically, recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} to understand
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

@@

\begin{figure*}[t]
\includegraphics[width=0.95\textwidth,clip,trim=0cm 6cm 1.7cm 0cm]{fig/walkthrough.pdf}
\caption{Organizing a conference using Denicek. The Walkthrough shows construction of a user
  interface for adding speakers (A, B, C); refactoring of the list and merging edits (D, E); and
  formulas with schema and code co-evolution (F, G).}
\label{fig:walkthrough}
\end{figure*}

@@

In this section, we use the web-based prototype Webnicek to illustrate programming
experiences enabled by the Denicek substrate. Webnicek is based on a structure
editor that supports navigation in the document and issuing of edit commands. We follow two
formative examples (\S\ref{app:examples}) where users collaboratively plan a conference.

\subsubsection*{\circled{A} Adding a Speaker}
The first user starts with an empty document, which is represented as a record. They add a field for
each heading and a field named {\small\Verb_speakers_} containing a list {\small\Verb_<ul>_}. They
use the command toolbox to add the first speaker.

\subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
further speakers, the user creates a textbox and a button. They enter new speaker
details, add a new {\small\Verb_<li>_} element to the list and copy the speaker details
from the textbox into the new element using a source view.

\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker from the textbox,
the user opens history view, selects edits that added the speaker and saves those as the
{\small\Verb_add-speaker_} interaction. They attach this interaction as a event handler for the
{\small\Verb_click_} event of the button.

@@

\subsubsection*{\circled{D} Refactoring Document Structure} Another user starts with the initial
version of the document (A) and turns the list into a table. This is done by invoking a series of
commands that change tags, wrap elements, copy values and transform them.

\subsubsection*{\circled{E} Merging Edits} The two document versions (C and D) can be automatically merged.
The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
button are also automatically transformed to the new format.

@@

\subsubsection*{\circled{F} Adding Budget Calculation} A third user adds a budget calculation to
the initial document. Formulas are represented as nodes with the {\small\Verb_x-formula_} tag,
whose arguments are other nodes, references to other nodes, or formulas (highlighted on mouse hover).

@@

\subsubsection*{\circled{G} Merging Formulas} When the budget calculation (F) is merged with the
earlier edits (E), references in formulas are automatically updated to point to the new list.
Adding new speaker via the ``Add speaker!'' button invalidates the evaluated result.

@@

@@

\begin{tabular}{|cl|}
\hline
\rowcolor{ekgray}
&\\[-1em]
\rowcolor{ekgray}
& \sffamily\small{\bfseries Kind}\;\,\textit{\footnotesize arguments} \\[0.2em]
\hline
&\\[-1em]
\ndtablecol{\faListUl}{List}{tag, index$_1$, child$_1$, $\ldots$, index$_n$, child$_n$}
  {Ordered list of nodes, addressable by \textit{index}. Renders as \Verb|<tag>| with children.}
\ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
  {Record with children addressable by \textit{field}. Renders as \Verb|<tag>| with children.}
\ndtablecol{\faExternalLink}{Reference}{selectors}
  {Reference to another document location. Displays the \Verb|selectors| as a link.}
\ndtablecol{\faFont}{Primitive}{string \textnormal{or} number}
  {Numerical or textual primitive value. Renders as an HTML text node.}
\hline
\end{tabular}
\vspace{-0.5em}
\caption{Structure of selectors and document nodes}
\label{fig:doc}
\vspace{-1em}
\end{figure}

\begin{figure}
\newcommand{\ektablecol}[5]{
\raisebox{-0.2em}{#1} & \sffamily\small{\bfseries #2}\,\;\textit{\footnotesize target, #3} & \sffamily\footnotesize #4 \\[-0.2em]
&\multicolumn{2}{l|}{\sffamily\footnotesize #5}\\[0.3em]
}
\begin{tabular}{|cp{16em}r|}
\hline
\rowcolor{ekgray}
&&\\[-1em]
\rowcolor{ekgray}
 & \sffamily\small{\bfseries Edit}\;\,\textit{\footnotesize arguments} & \sffamily\footnotesize\bfseries Target \\[0.2em]
\hline
&&\\[-1em]
\ektablecol{\faPlus}{Add}{field, after, node}{Record}
  {Add \textit{node} as a \textit{field} to the specified record \textit{after} a given field.}
\ektablecol{\faAt}{Append}{index, after, node}{List}
  {Append \textit{node} to the end of the specified list \textit{after} a given field.}
\ektablecol{\faSort}{Reorder}{permutation}{List}
  {Reorder items of a specified list according to a \textit{permutation}.}
\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}
  {Delete the item at a given \textit{index} of a specified list.}
\ektablecol{\faCode}{UpdateTag}{tag}{List or Record}
  {Change the tag of a specified list or record from to a new \textit{tag}.}
\ektablecol{\faICursor}{PrimitiveEdit}{transform}{Primitive}
  {Apply primitive \textit{transform} to the specified primitive.}
\hline
&&\\[-1em]
\ektablecol{\faFont}{RenameField}{old field, new field}{Record}
  {Rename the field of a specified record from \textit{old} to \textit{new}.}
\ektablecol{\faTimesCircle}{DeleteField}{field}{Record}
  {Delete the field \textit{field} of a specified record.}
\ektablecol{\faFileO}{WrapRecord}{tag, field}{Any}
  {Wrap the specified node as a \textit{field} of a new record with \textit{tag}.}
\ektablecol{\faListUl}{WrapList}{tag, index}{Any}
  {Wrap the specified node as a sole element of a new list with \textit{tag}.}
\ektablecol{\faCopy}{Copy}{selectors}{Any}
  {Copy nodes(s) from \textit{selectors}, replacing the specified target(s). }
\hline
\end{tabular}
\vspace{-0.75em}
\caption{Summary of document edit types in Denicek}
\label{fig:edits}
\vspace{-1em}
\end{figure}

@@

@@

\begin{figure}[t]
\includegraphics[width=0.95\columnwidth,clip,trim=1.7cm 6cm 7.8cm 1.5cm]{fig/merging.pdf}
\vspace{-0.5em}
\caption{Merging of two independently done sequences of edits. Two ways of merging B and C result in the same D.}
\label{fig:merging}
\vspace{-0.5em}
\end{figure}

@@

\subsubsection*{Conflict Resolution}
When merging two sequences of edits, $E, E_1$ and $E, E_2$, it is desirable that
applying $\mathcal{M}_E(E_1, E_2)$ and $\mathcal{M}_E(E_2, E_1)$ results in the same document.
This is not always the case. If the two edits modify the same value, transform the structure of
a node in incompatible ways or one edit modifies a node deleted by the other, the merge operation
is not symmetric. Such conflicting edits can be detected, reported to the user and optionally
removed. Conflict detection is also used to implement incremental evaluation~(\S\ref{sec:impl-incremental}).

@@

@@

Webnicek also accounts for the case where edits recorded in the document are themselves
transformed (when they are reconciliated with other edits during merging). In this case, Webnicek
updates the recorded edits (an alternative approach is discussed in \S\ref{sec:discuss}).

@@

@@

\keyideabox{\faMagic}{Generalization Heuristic}{Specifying generalization manually is cumbersome.
Programming by demonstration systems typically implement heuristic for generalization
\cite{myers-2000-intelligence} that infers and suggests such generalizations. If integrated into
Webnicek, such heuristic could for example automatically construct a formula
based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected Todo
list items).}

@@

\begin{figure}[t]
\vspace{-0.5em}
\includegraphics[width=1\columnwidth,clip,trim=0.5cm 8.5cm 8.5cm 0.5cm]{fig/interactive.pdf}
\vspace{-1.25em}
\caption{Using programming by demonstration to define a UI. The ``Add'' button (1) replays edits;
the ``Remove completed'' button (2) modifies target and specifies a condition. }
\label{fig:interactive}
\vspace{-1em}
\end{figure}

@@

\begin{figure}[t]
\vspace{-0.5em}
\includegraphics[width=0.95\columnwidth,clip,trim=0cm 5cm 11.5cm 0cm]{fig/counter.pdf}
\vspace{-0.5em}
\caption{We wrap the initial count (1) in a formula that adds 1 to the
value, (2). Evaluation produces the count (3), which is invalidated on subsequent clicks (4).
The button replays the saved edits that wrap the current count in a formula~(5).}
\label{fig:counter}
\vspace{-1em}
\end{figure}

@@

\begin{figure}[t]
  \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
  \vspace{-1em}
  \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
  is added (3), only the results of affected formulas are invalidated.}
  \label{fig:incremental}
  \vspace{-1em}
\end{figure}

@@

\begin{figure}[t]
\centering
\vspace{-0.5em}
\begin{minipage}{0.55\columnwidth}
  ~\\[1em]
  \includegraphics[width=0.9\columnwidth,clip,trim=0cm 14cm 21cm 0cm]{fig/eval.pdf}
  \end{minipage}%
  \begin{minipage}{0.45\columnwidth}
    \caption{Evaluated edits (A, B) are kept at the top of the history. Evaluated edits (A)
      that conflict with ordinary edits (4) are dropped.}
    \label{fig:eval}
  \end{minipage}
  \vspace{-1em}
\end{figure}

@@

@@

\begin{figure}[t]
\vspace{-0.25em}
\includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
\vspace{-2em}
\caption{Using provenance tracking to highlight document parts that contributed to the
calculation of refreshments costs (1), travel costs (2) and all costs (3).}
\label{fig:provenance}
\vspace{-0.75em}
\end{figure}

@@

@@

In Fig.~\ref{fig:copypaste}, the user copies a formula with switched operands using
the \ident{Copy} edit and then modifies it to use a different data source. They then navigate back
in history to the point before the copying and create a temporary fork of the document. In the
fork, they use \ident{RenameField} to switch the operands of the division. When they merge the
temporary fork into the original document, the \emph{Apply to Newly Added} logic of the merge
operation (\S\ref{sec:system-ops}) duplicates the \ident{RenameField} edits and applies them
to the copied formula. Merging with the \ident{Copy} edit and the fact that formulas
are ordinary document nodes, provide the key component for a straightforward implementation of the
managed copy \& paste functionality.

@@

\begin{figure}[t]
\vspace{-0.5em}
\includegraphics[width=0.9\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
\vspace{-1em}
\caption{When an edit changes the document structure, references in formulas are updated accordingly.}
\label{fig:coevolution}
\vspace{-0.5em}
\end{figure}

@@

\begin{figure}[t]
\vspace{-0.5em}
\includegraphics[width=1\columnwidth,clip,trim=0cm 5.5cm 1cm 0cm]{fig/copypaste.pdf}
\vspace{-1.5em}
\caption{The user uses creates a copy of a formula (1). They notice an error (2), go back in
history to switch the arguments (3), merge the change and re-evaluate both formulas (4).}
\label{fig:copypaste}
\vspace{-0.5em}
\end{figure}

@@

@@

\begin{figure}[t]
\vspace{-0.25em}
\includegraphics[width=0.9\columnwidth,clip,trim=0cm 0.7cm 18.5cm 0.2cm]{fig/source.pdf}
\vspace{-0.5em}
\caption{Underlying document representation of a code cell and a grid cell in the Datnicek
notebook system.}
\label{fig:source}
\vspace{-0.25em}
\end{figure}

\begin{figure*}[t]
\vspace{-0.5em}
\includegraphics[width=0.98\textwidth,clip,trim=0cm 1cm 1cm 0.5cm]{fig/datnicek.pdf}
\vspace{-0.5em}
\caption{A notebook visualizing Air traffic accident data from Eurostat. The user loads data in
a code cell (1) and edits it in a grid that infers edit operations via programming by demonstration
(2). They turn the edits into a code cell (3) and add a chart.}
\label{fig:datnicek}
\vspace{-0.5em}
\end{figure*}

@@

@@

The development of Datnicek was only started after the design and implementation of Denicek and
Webnicek was completed. Thanks to this two-phase methodology, the experience reported here
provides a qualitative evaluation of Denicek's capabilities. It also highlights limitations and
suggests directions for future work.

@@

Finally, data loaded in Datnicek is represented as Denicek nodes (a table is a list of
records). Such uniform representation makes De\-ni\-cek more versatile, but places a high demand on the
system performance. We implemented a few basic Denicek optimisations, but Datnicek is still cumbersome
to use with data consisting of multiple thousands of rows. Improving system performance is
thus an important direction for future work.

@@

Finally, the representation of transformations in interactive grid cells (Fig.~\ref{fig:source})
highlights an unnecessary duplication -- a transformation stores a sequence of edits, as well as a formula.
An interesting future direction is unifying or integrating those two notions of computation
provided by the Denicek substrate.

@@

\subsection{Technical Dimensions}
The technical dimensions (TDPS) framework \cite{jakubovic-2023-techdims} maps the design space of
interactive programming systems. We use the framework to characterize the generality of the
substrate.  For some dimensions, Denicek can cover the full range of options, while for others,
it is limited to one fixed design or a subset of options.

@@

@@

\newpage
~
\newpage

\appendix
\section{Formative Examples}
\label{app:examples}

>>>>

%\newcommand{\diff}[1]{{\textcolor{magenta}{#1}}}
%\newcommand{\note}[1]{\marginpar{\raggedright\scriptsize\textcolor{magenta}{#1}}}

\newcommand{\diff}[1]{{#1}}
\newcommand{\note}[1]{}

@@

\begin{teaserfigure}
  \vspace{-0.2em}
  \includegraphics[width=1\textwidth,clip,trim=0cm 8.5cm 7.7cm 0cm]{fig/teaser.pdf}
  \vspace{-1.5em}
  \caption{\diff{Denicek is a computational substrate for document-oriented programming
    based on document edit histories~(A). We co-design Denicek with a web-based end-user
    programming environment Denicek (B). Here, Denicek is used to build a Todo list app via
    programming-by-demonstration, by copying a value from input to a new list item.
    We then evaluate the generality of Denicek by using it to build a data science notebook
    system Datnicek (C). Here, Datnicek is used to interactively clean the data table, by
    removing the `` p'' (provisional) marker from numerical columns.
  }}
  \Description{The Denicek diagram in the middle illustrates a document representation as an
   edit history. Two arrows point from the diagram to two screenshots. Webnicek on the left
   contains a Todo list app with a command toolbox showing add item operation. Datnicek on the
   right contains two cells, one with code to load data and another with an interactive editor
   for data cleaning.}
  \label{fig:teaser}
  \vspace{0.75em}
\end{teaserfigure}
\maketitle

@@

\begin{itemize}
\item \emph{Collaborative Editing.} Users should be able to locally modify a shared document
  and merge their changes, preferably in a way that does not require a live connection to a central server~\cite{kleppmann-2019-local}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  enacting the steps of the desired behavior using concrete examples and generalizing from the example~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on the part are invalidated and, possibly automatically, recomputed \cite{teitelbaum-1981-cps,mcdirmid-2013-usable,horowitz-2023-engraft}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} to understand
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

@@

\begin{figure*}[t]
\includegraphics[width=0.95\textwidth,clip,trim=0cm 6cm 1.7cm 0cm]{fig/walkthrough.pdf}
\vspace{-0.5em}
\caption{Organizing a conference using Denicek. The Walkthrough shows construction of a user
  interface for adding speakers (A, B, C); refactoring of the list and merging edits (D, E); and
  formulas with schema and code co-evolution (F, G).}
\label{fig:walkthrough}
\end{figure*}

@@

We use the web-based prototype Webnicek to illustrate programming experiences enabled by Denicek.
Webnicek is based on a structure editor that supports navigation in the document and issuing of
edit commands. \diff{We follow a formative example (\S\ref{app:examples}) where Evelyn, Juliana and Alfred
collaboratively plan a conference.}
\note{Turn this into a concrete user story consistent with the video. Also add more details about
the edit operations and UI interactions.}

\subsubsection*{\circled{A} Adding a Speaker}
Evelyn starts with an empty document, which is represented as a record. She adds a field for
each heading and a field named {\small\Verb_speakers_} containing a list {\small\Verb_<ul>_}. \diff{She
uses the command toolbox to issue edit commands that add the first speaker.}

\subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
further speakers, Evelyn creates a textbox and a button \diff{using the command toolbox}. She enters
the details of another speaker \diff{into the textbox}, adds a new {\small\Verb_<li>_} element
and copies the speaker details from the textbox into the new element in the source view,
\diff{using a copy command}.

\subsubsection*{\circled{C} Abstracting the Interaction} After adding the speaker,
Evelyn opens the history view, selects edits that added the speaker and saves them as the
{\small\Verb_add-speaker_} interaction. She attaches this interaction as an event handler for the
{\small\Verb_click_} event of the button.

@@

\subsubsection*{\circled{D} Refactoring Document Structure} Alfred starts with the initial
version of the document (A) and turns the list into a table. He invokes a series of commands that
\diff{rename tags, wrap elements, copy values and split strings using the comma as a separator}.
\note{More concrete description of edits}

\subsubsection*{\circled{E} Merging Edits}
\diff{Alfred then merges the later edits (B, C) done by Evelyn into his version of the document.}
The refactoring is applied to all speakers. New speakers added using the ``Add speaker!''
button are also automatically transformed to the new format.

@@

\subsubsection*{\circled{F} Adding Budget Calculation} Juliana joins in and adds a budget calculation to
the initial document. \diff{She uses the command toolbox to create document nodes representing
formulas. Formulas are regular nodes with a special} {\small\Verb_x-formula_} \diff{tag and
arguments as child nodes. To count the speakers, she uses the} {\small\Verb_count_} \diff{builtin
with a reference to the node representing the speaker list as the argument.}\note{Modified to describe
representation of formulas.}

@@

\subsubsection*{\circled{G} Merging Formulas} When Juliana merges the budget calculation (F) with
the earlier edits (E), references in formulas are automatically updated to point to the \diff{rows of the table}.
Adding new speaker via the ``Add speaker!'' button \diff{invalidates results of only those formulas
that depend on the number of speakers.}

@@

\begin{tabular}{|cl|}
\hline
\rowcolor{ekgray}
&\\[-1em]
\rowcolor{ekgray}
& \sffamily\small{\bfseries Kind}\;\,\textit{\footnotesize arguments} \\[0.2em]
\hline
&\\[-1em]
\ndtablecol{\faListUl}{List}{tag, index$_1$, child$_1$, $\ldots$, index$_n$, child$_n$}
  {Ordered list of nodes, addressable by \textit{index}. Renders as \Verb|<tag>| with children.}
\ndtablecol{\faFileO}{Record}{tag, field$_1$, child$_1$, $\ldots$, field$_n$, child$_n$}
  {Record with children addressable by \textit{field}. Renders as \Verb|<tag>| with children.}
\ndtablecol{\faExternalLink}{Reference}{selectors}
  {Reference to another document location. Displays the \Verb|selectors| as a link.}
\ndtablecol{\faFont}{Primitive}{string \textnormal{or} number}
  {Numerical or textual primitive value. Renders as an HTML text node.}
\hline
\end{tabular}
\caption{Structure of selectors and document nodes}
\label{fig:doc}
\end{figure}

@@

@@

\begin{figure}
\newcommand{\ektablecol}[5]{
\raisebox{-0.2em}{#1} & \sffamily\small{\bfseries #2}\,\;\textit{\footnotesize target, #3} & \sffamily\footnotesize #4 \\[-0.2em]
&\multicolumn{2}{l|}{\sffamily\footnotesize #5}\\[0.3em]
}
\begin{tabular}{|cp{16em}r|}
\hline
\rowcolor{ekgray}
&&\\[-1em]
\rowcolor{ekgray}
 & \sffamily\small{\bfseries Edit}\;\,\textit{\footnotesize arguments} & \sffamily\footnotesize\bfseries Target \\[0.2em]
\hline
&&\\[-1em]
\ektablecol{\faPlus}{Add}{field, after, node}{Record}
  {Add \textit{node} as a \textit{field} to the specified record \textit{after} a given field.}
\ektablecol{\faAt}{Append}{index, after, node}{List}
  {Append \textit{node} to the end of the specified list \textit{after} a given field.}
\ektablecol{\faSort}{Reorder}{permutation}{List}
  {Reorder items of a specified list according to a \textit{permutation}.}
\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}
  {Delete the item at a given \textit{index} of a specified list.}
\ektablecol{\faCode}{UpdateTag}{tag}{List or Record}
  {Change the tag of a specified list or record from to a new \textit{tag}.}
\ektablecol{\faICursor}{PrimitiveEdit}{transform}{Primitive}
  {Apply primitive \textit{transform} to the specified primitive.}
\hline
&&\\[-1em]
\ektablecol{\faFont}{RenameField}{old field, new field}{Record}
  {Rename the field of a specified record from \textit{old} to \textit{new}.}
\ektablecol{\faTimesCircle}{DeleteField}{field}{Record}
  {Delete the field \textit{field} of a specified record.}
\ektablecol{\faFileO}{WrapRecord}{tag, field}{Any}
  {Wrap the specified node as a \textit{field} of a new record with \textit{tag}.}
\ektablecol{\faListUl}{WrapList}{tag, index}{Any}
  {Wrap the specified node as a sole element of a new list with \textit{tag}.}
\ektablecol{\faCopy}{Copy}{selectors}{Any}
  {Copy nodes(s) from \textit{selectors}, replacing the specified target(s). }
\hline
\end{tabular}
\caption{Summary of document edit types in Denicek}
\label{fig:edits}
\end{figure}

@@

@@

\subsubsection*{Conflict Resolution}
When merging two sequences of edits, $E, E_1$ and $E, E_2$, it is desirable that
$\mathcal{M}_E(E_1, E_2)$ and $\mathcal{M}_E(E_2, E_1)$ result in the same document.
This is not always the case. If the two edits modify the same value, transform the structure of a node
in incompatible ways or one edit modifies a node deleted by the other, the merge operation
is not symmetric. Such conflicting edits can be detected and reported to the user.
\diff{Conflict detection is also used to implement incremental recomputation~(\S\ref{sec:impl-incremental}),
in which case conflicting edits produced by evaluation are removed and formulas have to be
recomputed.}\note{Clarify conflicting edits are only removed in incremental evaluation}

@@

\begin{figure}[t]
\includegraphics[width=0.95\columnwidth,clip,trim=1.7cm 6cm 7.8cm 1.5cm]{fig/merging.pdf}
\caption{Merging of two independently done sequences of edits. Two ways of merging B and C result in the same D.}
\label{fig:merging}
\end{figure}

@@

Webnicek also accounts for the case where edits recorded in the document are themselves
transformed (when they are reconciled with other edits during merging). In this case, Webnicek
updates the recorded edits (an alternative approach is discussed in \S\ref{sec:discuss}).

@@

\begin{figure}[t]
\includegraphics[width=0.95\columnwidth,clip,trim=0cm 5cm 11.5cm 0cm]{fig/counter.pdf}
\caption{We wrap the initial count (1) in a formula that adds 1 to the
value, (2). Evaluation produces the count (3), which is invalidated on subsequent clicks (4).
The button replays the saved edits that wrap the current count in a formula~(5).}
\label{fig:counter}
\end{figure}

@@

\keyideabox{\faMagic}{Generalization Heuristic}{Specifying generalization manually is cumbersome.
Programming by demonstration systems typically implement heuristics for generalization
\cite{myers-2000-intelligence} that infers and suggests such generalizations. If integrated into
Webnicek, such heuristic could for example automatically construct a formula
based on positive and negative examples \cite{gulwani-2014-flash} (selected and deselected Todo
list items).}

@@

\begin{figure}[t]
\includegraphics[width=1\columnwidth,clip,trim=0.5cm 8.5cm 8.5cm 0.5cm]{fig/interactive.pdf}
\caption{Using programming by demonstration to define a UI. The ``Add'' button (1) replays edits;
the ``Remove completed'' button (2) modifies target and specifies a condition. }
\label{fig:interactive}
\end{figure}

@@

@@

\begin{figure}[t]
  \includegraphics[width=1\columnwidth,clip,trim=0.1cm 5cm 5.1cm 0cm]{fig/incremental.pdf}
  \caption{Budget calculation based on the number of speakers (1) and the result (2). When speaker
  is added (3), only the results of affected formulas are invalidated.}
  \label{fig:incremental}
  \vspace{-1em}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{minipage}{0.55\columnwidth}
  ~\\[1em]
  \includegraphics[width=0.9\columnwidth,clip,trim=0cm 14cm 21cm 0cm]{fig/eval.pdf}
  \end{minipage}%
  \begin{minipage}{0.45\columnwidth}
    \caption{Evaluated edits (A, B) are kept at the top of the history. Evaluated edits (A)
      that conflict with ordinary edits (4) are dropped.}
    \label{fig:eval}
  \end{minipage}
\end{figure}

@@

\diff{Note that Denicek can only reflect schema changes explicitly represented by the document
structure. If a formula depended on the structure of the original string representing speakers
(with a comma), Denicek would not be able to amend the logic of the formula. Systems supporting
the divergence model \cite{edwards-2025-schema} may be able to lift this restriction by
transforming the new value back before applying the original formula.}
\note{Added response to R3's question about schema change}

@@

\begin{figure}[t]
\includegraphics[width=1\columnwidth,clip,trim=0cm 8cm 5cm 0cm]{fig/provenance.pdf}
\vspace{-1em}
\caption{Using provenance tracking to highlight document parts that contributed to the
calculation of refreshments costs (1), travel costs (2) and all costs (3).}
\label{fig:provenance}
\end{figure}

@@

\begin{figure}[t]
\includegraphics[width=0.9\columnwidth,clip,trim=0cm 9.5cm 8cm 0cm]{fig/coevolution.pdf}
\caption{When an edit changes the document structure, references in formulas are updated accordingly.}
\label{fig:coevolution}
\end{figure}

@@

\note{Modified figure to add call to read-csv}
\diff{The functionality is illustrated in Fig.~\ref{fig:copypaste}, which uses the builtin operation}
{\small\Verb_read-csv_} \diff{to load a table from a file. The operation imports the data as a
document table with rows that can be addressed by subsequent formulas. The then user writes two
formulas to sum the rows and compute an average.}
\note{Add explanation of a more complex builtin operation}

They copy a formula with switched operands using
the \ident{Copy} edit and then modify it to use a different data source. They then navigate back
in history to the point before the copying and create a temporary fork of the document. In the
fork, they use \ident{RenameField} to switch the operands of the division. When they merge the
temporary fork into the original document, the \emph{Apply to Newly Added} logic of the merge
operation (\S\ref{sec:system-ops}) duplicates the \ident{RenameField} edits and applies them
to the copied formula. Merging with the \ident{Copy} edit and the fact that formulas
are ordinary document nodes, provide the key component for a straightforward implementation of the
managed copy \& paste functionality.

@@

\begin{figure}[t]
\includegraphics[width=1\columnwidth,clip,trim=0cm 1.7cm 1cm 0cm]{fig/copypaste.pdf}
\caption{The user uses creates a copy of a formula (1). They notice an error (2), go back in
history to switch the arguments (3), merge the change and re-evaluate both formulas (4).}
\label{fig:copypaste}
\end{figure}

@@

@@

\note{Respond to R3's question about avoiding edits depending on values and R1's concern about
usefulness for PBD tools.}
\diff{This limitation means that edits generated by generalized interactions (\S\ref{sec:impl-interaction})
can be only merged in limited ways (an item added to a list through merging after
invoking an action will not be affected by the action). We expect that the restriction on
expressiveness of selectors can be lifted for edits that do not modify the document structure,
but have not yet lifted this restriction in Denicek.}

@@

\begin{figure*}[t]
\vspace{-0.5em}
\includegraphics[width=0.98\textwidth,clip,trim=0cm 1cm 1cm 0.5cm]{fig/datnicek.pdf}
\vspace{-0.5em}
\caption{A notebook visualizing Air traffic accident data from Eurostat. The user loads data in
a code cell (1) and edits it in a grid that infers edit operations via programming by demonstration
(2). They turn the edits into a code cell (3) and add a chart.}
\label{fig:datnicek}
\end{figure*}

@@

\begin{figure}[t]
\includegraphics[width=0.9\columnwidth,clip,trim=0cm 0.7cm 18.5cm 0.2cm]{fig/source.pdf}
\caption{Underlying document representation of a code cell and a grid cell in the Datnicek
notebook system.}
\label{fig:source}
\end{figure}

@@

\subsubsection*{External Validity.}
The development of Datnicek was only started after the design and implementation of Denicek and
Webnicek was completed. Thanks to this two-phase methodology, the case study provides
a qualitative evaluation of Denicek's capabilities. \diff{The design of Denicek is based
on third-party research (\S\ref{sec:case-req}) and so the case study provides a limited external
validation. However, several design choices of Datnicek align it with the capabilities of Denicek.
This includes the use of a structure editor and
a user interface for programming by demonstration that suggests edit operations. We revisit the
generality of Denicek in \S\ref{sec:eval-tdps}, but establishing external validity ultimately
requires adoption of the open-source Denicek package by other programming system researchers.}
\note{Added discussion on external validity - although this is necessarilly limited.}

@@

Finally, data loaded in Datnicek is represented as Denicek nodes (a table is a list of
records). Such uniform representation makes De\-ni\-cek more versatile, but places a high demand on the
system performance. We implemented a few basic Denicek optimisations, but Datnicek is still cumbersome
to use with data consisting of multiple thousands of rows.

@@

Finally, the representation of transformations in interactive grid cells (Fig.~\ref{fig:source})
highlights an unnecessary duplication -- a transformation stores a sequence of edits, as well as a formula.
Unifying these two notions is another interesting future direction.

@@

\subsection{Technical Dimensions}
\label{sec:eval-tdps}

The technical dimensions (TDPS) framework \cite{jakubovic-2023-techdims} maps the design space of
interactive programming systems. We use the framework to characterize the generality of the
substrate.  For some dimensions, Denicek can cover the full range of options, while for others,
it is limited to one fixed design or a subset of options.

@@

\begin{acks}
The authors thank Clemens Nylandsted Klokmose for research guidance and
attendees of the IFIP WG 2.16 meeting in Serpiano, members of the Software Architecture Group at HPI
and attendees of LIVE 2023 for feedback on the project.
Anonymous reviewers provided invaluable actionable guidance for improving the final version
of the paper. The work has been supported by the
\grantsponsor{cuni}{Charles University}{https://d3s.mff.cuni.cz/projects/primus24/}
grant \grantnum{cuni}{PRIMUS/24/SCI/021} and by the
\grantsponsor{erccz}{Czech Ministry of Education, Youth and Sports}{https://starfos.tacr.cz/en/projekty/LL2325}
grant ERC-CZ \grantnum{erccz}{LL2325}.
\end{acks}

@@

\appendix
